How Kids Code and How We Know:\\An Exploratory Study on the Scratch Repository

### 2016-04-22 Paper/paper.tex
Fixed table 3

\newcommand{\nPrograms}{250,166}
\newcommand{\nAnalyzedPrograms}{247,798}
\newcommand{\nemptyPrograms}{14,307}
\newcommand{\nScriptPrograms}{233,491}
\newcommand{\nProgramsWithClones}{67,177}
\newcommand{\nLOC}{36,085,654}
\newcommand{\nscripts}{4,049,356}
\newcommand{\dup}{\emph{Duplication}~}

@@

Once we obtained the Scratch projects, we parsed the JSON files according to the specification of the format.\footnote{\url{http://wiki.scratch.mit.edu/wiki/Scratch_File_Format_(2.0)}} This resulted in a list of used blocks per project, with the sprites and the stage of the project. We cross referenced also all blocks with the Scratch wiki so we have the shapes and the category of all blocks. For example, When Green Flag Clicked is a \emph{Hat block} from the \emph{Events category}.

@@

\subsection{Program Size and Complexity}
\label{RQ1}
The dataset contains a relatively small number of projects without any code: \nemptyPrograms (5.77\%). Through random manual sampling we found that in some cases these projects contains only sprites and costumes, but no code, while in other projects they were entirety empty apart from the Scratch cat added by default. Since these projects are empty in terms of code we excluded them from further analysis, leaving the final number of analyzed non-empty projects to \nScriptPrograms.

\begin{table*}[ht]
	\centering
	\begin{tabular}{lrrrrrr}
		\hline
		&\textbf{mean}&\textbf{min}&\textbf{Q1}&\textbf{median}&\textbf{Q3}&\textbf{max}\\
		\hline
		Sprites with code per project&5.68&1&1&2&5&525\\
		Scripts per project&17.35&1&2&5&12&3038\\
		Number of blocks per project&154.55&1&12&29&76&34622\\
		Blocks in Stage per project&4.80&0&0&0&3&2613\\
		Blocks in Sprites per project&115.57&0&10&26&68&34613\\
		Blocks in Procedures per project&34.17&0&0&0&0&20552\\
		McCabe Cyclomatic Complexity (CC) per script&1.58&1&1&1&1&246\\
		McCabe CC per procedure script&3.75&1&1&2&4&183\\
		Procedures per project with procedures&11.50&1&1&2&6&847\\
		Arguments per Procedure&0.95&0&0&0&1&53\\
		Numerical arguments per procedure with arguments&1.73&0&1&1&2&22\\
		Text arguments per procedure with arguments&0.28&0&0&0&1&24\\
		Boolean arguments per procedure with arguments&0.13&0&0&0&0&14\\
		Calls per procedure&2.14&0&1&1&2&526\\
		Scripts with calls per procedure&1.13&0&1&1&1&59\\
		Variables per project&2.06&0&0&0&1&340\\
		Scripts utilizing variable&4.97&1&1&3&5&1127\\
		Lists per project&0.55&0&0&0&0&319\\
		Conditional statements per project&10.02&0&0&0&3&5950\\
		Loop statements per project&7.65&0&1&2&5&2503\\
		User input blocks per project&4.77&0&0&1&4&1889\\
		Broadcast-receive statements per project&8.57&0&0&0&2&2460\\	
		\hline
	\end{tabular}
	\caption{Summary statistics from the dataset of \nScriptPrograms~non-empty Scratch projects}
	\label{tbl-size}
\end{table*}

@@

We further analyzed the utilization frequency of the different block shapes and categories, as defined in the Scratch documentation. Figures \ref{fig:categories} and \ref{fig:shapes} present the results in terms of number of blocks from the total \nLOC blocks in the dataset projects. The most commonly used blocks are from the Control and Data categories. The Others category includes the blocks related to procedure calls and arguments.

@@

\begin{table}
	\centering
	\begin{tabular}{lrr}
		\hline
		\textbf{Block}&\textbf{Projects}&\textbf{Occurrences}\\
		\hline
	\texttt{when <> key pressed}&71,096&294,771\\
	\texttt{when this sprite clicked}&39,179&198,342\\
	(Sensing) \texttt{key <> pressed?}&37,919&291,657\\
	(Sensing) \texttt{ask <> and wait}&19,039&66,850\\
	(Sensing) \texttt{mouse down?}&9,115&54,079\\
	(Sensing) \texttt{<attrib> of <>}&9,068&155,468\\
	(Sensing) \texttt{mouse X}&5,977&27,321\\
	(Sensing) \texttt{mouse Y}&3,940&22,035\\
	\texttt{when <sensor> > <value>}&705&1,570\\
	(Sensing) \texttt{video <> on <>}&434&1,397\\
		\hline
	\end{tabular}
	\caption{Frequency of use of user input blocks in the \nScriptPrograms projects of the dataset}
	\label{tbl-userinput}
\end{table}

@@

\section{Conclusions}
\label{sec:conclusion}
In this paper we presented a large scale study on \nAnalyzedPrograms projects we scraped from the Scratch repository. We analyzed these projects in terms of size, complexity, application of programming abstractions and utilization of programming concepts including procedures, variables, conditional statements, loops, and broadcast-receive functionality. We found that procedures and conditional loops are not commonly used. We further investigated the presence of code smells, including code duplication, dead code, long method and large class smells. Our findings indicate that Scratch programs suffer from code smells and especially dead code and code duplication.

>>>>

\newcommand{\nPrograms}{250,166}
\newcommand{\nAnalyzedPrograms}{247,798}
\newcommand{\nemptyPrograms}{14,307}
\newcommand{\nScriptPrograms}{233,491}
\newcommand{\nLOC}{36,085,654}
\newcommand{\nscripts}{4,049,356}

@@

Once we obtained the Scratch projects, we parsed the JSON files according to the specification of the format.\footnote{\url{http://wiki.scratch.mit.edu/wiki/Scratch_File_Format_(2.0)}} This resulted in a list of used blocks per project, with the sprites and the stage of the project. We cross referenced also all blocks with the Scratch wiki so we have the shapes and the category of all blocks. For example, \texttt{When Green Flag Clicked} is a \emph{Hat block} from the \emph{Events category}.

@@

\subsection{Program Size and Complexity}
\label{RQ1}
The dataset contains a relatively small number of projects without any code: \nemptyPrograms~ (5.77\%). Through random manual sampling we found that in some cases these projects contains only sprites and costumes, but no code, while in other projects they were entirety empty apart from the Scratch cat added by default. Since these projects are empty in terms of code we excluded them from further analysis, leaving the final number of analyzed non-empty projects to \nScriptPrograms.

\begin{table*}[ht]
	\centering
	\begin{tabular}{llrrrrrr}
		&&\textbf{mean}&\textbf{min}&\textbf{Q1}&\textbf{median}&\textbf{Q3}&\textbf{max}\\
		\hline
		Size&Sprites with code per project&5.68&1&1&2&5&525\\
		&Scripts per project&17.35&1&2&5&12&3,038\\
		&Number of blocks per project&154.55&1&12&29&76&34,622\\
		&Blocks in Stage per project&4.80&0&0&0&3&2,613\\
		&Blocks in Sprites per project&115.57&0&10&26&68&34,613\\
		&Blocks in Procedures per project&34.17&0&0&0&0&20,552\\
		\hline
		Complexity&McCabe Cyclomatic Complexity (CC) per script&1.58&1&1&1&1&246\\
		&McCabe CC per procedure script&3.75&1&1&2&4&183\\
		\hline
		Procedures&Procedures per project with procedures&11.50&1&1&2&6&847\\
		&Arguments per Procedure&0.95&0&0&0&1&53\\
		&Numerical arguments per procedure with arguments&1.73&0&1&1&2&22\\
		&Text arguments per procedure with arguments&0.28&0&0&0&1&24\\
		&Boolean arguments per procedure with arguments&0.13&0&0&0&0&14\\
		&Calls per procedure&2.14&0&1&1&2&526\\
		&Scripts with calls per procedure&1.13&0&1&1&1&59\\
		\hline
		Programming&Variables per project&2.06&0&0&0&1&340\\
		concepts&Scripts utilizing variable&4.97&1&1&3&5&1,127\\
		&Lists per project&0.55&0&0&0&0&319\\
		&Conditional statements per project&10.02&0&0&0&3&5,950\\
		&Loop statements per project&7.65&0&1&2&5&2,503\\
		&User input blocks per project&4.77&0&0&1&4&1,889\\
		&Broadcast-receive statements per project&8.57&0&0&0&2&2,460\\	
		\hline
	\end{tabular}
	\caption{Summary statistics from the dataset of \nScriptPrograms~non-empty Scratch projects}
	\label{tbl-size}
\end{table*}

@@

We further analyzed the utilization frequency of the different block shapes and categories, as defined in the Scratch documentation. Figures \ref{fig:categories} and \ref{fig:shapes} present the results in terms of number of blocks from the total \nLOC~blocks in the dataset projects. The most commonly used blocks are from the Control and Data categories. The Others category includes the blocks related to procedure calls and arguments.

@@

\begin{table}
	\centering
	\begin{tabular}{lrr}
		\hline
		\textbf{Block}&\textbf{Projects}&\textbf{Occurrences}\\
		\hline
	\texttt{when <> key pressed}&71,096&294,771\\
	\texttt{when this sprite clicked}&39,179&198,342\\
	(Sensing) \texttt{key <> pressed?}&37,919&291,657\\
	(Sensing) \texttt{ask <> and wait}&19,039&66,850\\
	(Sensing) \texttt{mouse down?}&9,115&54,079\\
	(Sensing) \texttt{<attrib> of <>}&9,068&155,468\\
	(Sensing) \texttt{mouse X}&5,977&27,321\\
	(Sensing) \texttt{mouse Y}&3,940&22,035\\
	\texttt{when <sensor> > <value>}&705&1,570\\
	(Sensing) \texttt{video <> on <>}&434&1,397\\
		\hline
	\end{tabular}
	\caption{Frequency of use of user input blocks in the \nScriptPrograms~projects of the dataset}
	\label{tbl-userinput}
\end{table}

@@

\section{Conclusions}
\label{sec:conclusion}
In this paper we presented a large scale study on \nAnalyzedPrograms~projects we scraped from the Scratch repository. We analyzed these projects in terms of size, complexity, application of programming abstractions and utilization of programming concepts including procedures, variables, conditional statements, loops, and broadcast-receive functionality. We found that procedures and conditional loops are not commonly used. We further investigated the presence of code smells, including code duplication, dead code, long method and large class smells. Our findings indicate that Scratch programs suffer from code smells and especially dead code and code duplication.
